---
title: "Student Depression Data Analysis Project"
output:
  pdf_document: default
  html_document: default
date: "2025-10-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidyverse)
library(ggplot2)
```

# Research questions:
1. Is there a correlation between the cumulative grade point average of the student and depression?
2. Which personal, academic, and lifestyle factors are associated with depression among students?
3. Is age associated with depression?
4. Is academic pressure associated with depression?
5. Does study satisfaction moderate the effect of academic pressure on depression?
6. Are students with higher work/study hours more likely to experience depression?
7. Do dietary habits (e.g., healthy vs unhealthy) relate to depression?
8. Does financial stress correlate with depression?

# Loading data and initial data exploration

```{r}
df <- read.csv("./dataset/student_depression_dataset.csv")
```

```{r}
head(df)
```
```{r}
names(df)
```

The data set consists of 27901 observations and 18 features
```{r}
dim(df)
```

```{r}
unique(df$Profession)
```

Convert to data.table
```{r}
dt <- as.data.table(df)
```

```{r}
dt[Profession != "Student"]
```
Count observations by profession
```{r}
dt[, .N, by="Profession"]
```
Count observations by city
There are observations of students from 52 cities.
```{r}
dt[, .N, by="City"][order(-N)]
```
Explore categories of "Sleep.Duration"
```{r}
unique(dt$Sleep.Duration)
```
Explore categories of "Dietary.Habits"
```{r}
unique(dt$Dietary.Habits)
```
Explore categories of "Degree"
```{r}
unique(dt$Degree)
```
Categories of "Have.you.ever.had.suicidal.thoughts.."
```{r}
unique(dt$Have.you.ever.had.suicidal.thoughts..)
```
Categories of "Financial.Stress"
```{r}
unique(dt$Financial.Stress)
```
Categories of "Family.History.of.Mental.Illness"
```{r}
unique(dt$Family.History.of.Mental.Illness)
```
Distribution of age
```{r}
ggplot(dt, aes(x = Age)) +
  geom_histogram(fill="pink", color="grey") + 
  ylab("Number of students") + 
  xlab("Age") + 
  ggtitle("Distribution of Participant Ages")
```
```{r}
ggplot(dt, aes(x = Gender)) +
  geom_bar(fill="lightblue", color="grey") + 
  ylab("Number of students") + 
  xlab("") +
  ggtitle("Distribution of gender among participants")
```
```{r}
ggplot(dt, aes(x = factor(Depression, levels=c(0, 1), labels=c("No", "Yes")))) +
  geom_bar(fill="lightgreen", color="grey") + 
  ylab("Number of students") + 
  xlab("") +
  ggtitle("Distribution of depressive vs. non-depressive participants")
```

```{r}
ggplot(dt, aes(x = Have.you.ever.had.suicidal.thoughts..)) +
  geom_bar(fill="lightblue", color="grey") + 
  ylab("Number of students") + 
  xlab("") +
  ggtitle("Distribution of students with/without suicidal thoughts")
```
# Data preprocessing
```{r}
head(dt)
```
Select variables of interest
```{r}
dt <- dt[, .(id, Gender, Academic.Pressure, CGPA, Study.Satisfaction, Sleep.Duration, Dietary.Habits, Have.you.ever.had.suicidal.thoughts.., Work.Study.Hours, Financial.Stress, Family.History.of.Mental.Illness, Depression)]
```

```{r}
dt$Gender <- as.factor(dt$Gender)
dt$Sleep.Duration <- as.factor(dt$Sleep.Duration)
dt$Dietary.Habits <- as.factor(dt$Dietary.Habits)
dt$Have.you.ever.had.suicidal.thoughts.. <- as.factor(dt$Have.you.ever.had.suicidal.thoughts..)
dt$Financial.Stress <- as.numeric(dt$Financial.Stress)
dt$Family.History.of.Mental.Illness <- as.factor(dt$Family.History.of.Mental.Illness)
```

```{r}
setnames(dt, old="Family.History.of.Mental.Illness", new="Family.Mental.Illness")
setnames(dt, old="Have.you.ever.had.suicidal.thoughts..", new="Suicidal.thoughts")
```

```{r}
head(dt)
```
Check for missing values
```{r}
colSums(is.na(df))
```
```{r}
str(dt)
```
```{r}
summary(dt)
```
# Answer research questions

Examine correlation between numeric variables

There is a significant positive correlation between academic pressure and depression (0.47).
There is a significant negative correlation between age and depression (-0.22).
```{r}
cols <- sapply(dt, is.numeric)
cor(dt[, ..cols], use = "complete.obs")
```
# Summary
