---
title: "Student Depression Data Analysis Project"
output:
  html_document: default
date: "2025-10-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidyverse)
library(ggplot2)
library(plotly)
```

# Research questions:
1. Is there a correlation between the cumulative grade point average of the student and depression?
2. Is age associated with depression?
3. Is academic pressure associated with depression?
4. Are students with higher work/study hours more likely to experience depression?
5. Do dietary habits (e.g., healthy vs unhealthy) relate to depression?
6. Does financial stress correlate with depression?

# Loading data and initial data exploration

```{r}
df <- read.csv("./dataset/student_depression_dataset.csv")
```

```{r}
head(df)
```
```{r}
names(df)
```

The data set consists of 27901 observations and 18 features
```{r}
dim(df)
```

```{r}
unique(df$Profession)
```

Convert to data.table
```{r}
dt <- as.data.table(df)
```

```{r}
dt[Profession != "Student"]
```
Count observations by profession
```{r}
dt[, .N, by="Profession"]
```
Count observations by city
There are observations of students from 52 cities.
```{r}
dt[, .N, by="City"][order(-N)]
```
Explore categories of "Sleep.Duration"
```{r}
unique(dt$Sleep.Duration)
```
Explore categories of "Dietary.Habits"
```{r}
unique(dt$Dietary.Habits)
```
Explore categories of "Degree"
```{r}
unique(dt$Degree)
```
Categories of "Have.you.ever.had.suicidal.thoughts.."
```{r}
unique(dt$Have.you.ever.had.suicidal.thoughts..)
```
Categories of "Financial.Stress"
```{r}
unique(dt$Financial.Stress)
```
Categories of "Family.History.of.Mental.Illness"
```{r}
unique(dt$Family.History.of.Mental.Illness)
```
Distribution of age
```{r}
p <- ggplot(dt, aes(x = Age)) +
  geom_histogram(fill="pink", color="grey") + 
  ylab("Number of students") + 
  xlab("Age") + 
  ggtitle("Distribution of Participant Ages")

ggplotly(p)
```
```{r}
p <- ggplot(dt, aes(x = Gender)) +
  geom_bar(fill="lightblue", color="grey") + 
  ylab("Number of students") + 
  xlab("") +
  ggtitle("Distribution of gender among participants")

ggplotly(p)
```
```{r}
p <- ggplot(dt, aes(x = factor(Depression, levels=c(0, 1), labels=c("No", "Yes")))) +
  geom_bar(fill="lightgreen", color="grey") + 
  ylab("Number of students") + 
  xlab("") +
  ggtitle("Distribution of depressive vs. non-depressive participants")

ggplotly(p)
```

```{r}
p <- ggplot(dt, aes(x = Have.you.ever.had.suicidal.thoughts..)) +
  geom_bar(fill="lightblue", color="grey") + 
  ylab("Number of students") + 
  xlab("") +
  ggtitle("Distribution of students with/without suicidal thoughts")

ggplotly(p)
```
# Data preprocessing
```{r}
head(dt)
```

Select variables of interest
```{r}
dt <- dt[, .(id, Age, Gender, Academic.Pressure, CGPA, Study.Satisfaction, Sleep.Duration, Dietary.Habits, Have.you.ever.had.suicidal.thoughts.., Work.Study.Hours, Financial.Stress, Family.History.of.Mental.Illness, Depression)]
```

Map boolean values to numerical
```{r}
dt[, Have.you.ever.had.suicidal.thoughts.. := ifelse(Have.you.ever.had.suicidal.thoughts.. == "Yes", 1L, 0L)]
```

```{r}
dt[, Family.History.of.Mental.Illness := ifelse(Family.History.of.Mental.Illness == "Yes", 1L, 0L)]
```

```{r}
dt$Gender <- as.factor(dt$Gender)
dt$Sleep.Duration <- as.factor(dt$Sleep.Duration)
dt$Dietary.Habits <- as.factor(dt$Dietary.Habits)
dt$Financial.Stress <- as.numeric(dt$Financial.Stress)
```

```{r}
setnames(dt, old="Family.History.of.Mental.Illness", new="Family.Mental.Illness")
setnames(dt, old="Have.you.ever.had.suicidal.thoughts..", new="Suicidal.thoughts")
```

```{r}
head(dt)
```


Check for missing values
```{r}
colSums(is.na(df))
```
```{r}
str(dt)
```
```{r}
summary(dt)
```
# Answer research questions

## Is there a correlation between the cumulative grade point average of the student and depression?

The following analysis illustrates that there is no statistically significant correlation between those two features.

```{r}
cor(dt$CGPA, dt$Depression)
```
```{r}
p <- ggplot(dt, aes(x=factor(Depression), y=CGPA, fill=Depression)) + 
  geom_boxplot(alpha=0.7) + 
  labs(
    title="Correlation between the cumulative grade point average and depression",
    x="Depression",
    y="CGPA"
  )

ggplotly(p)
```

## Is age associated with depression?

There is a significant negative correlation between age and depression (-0.22), indicating that younger participants are slightly more likely to experience depression.

```{r}
cor(dt$Age, dt$Depression)
```

```{r}
p <- ggplot(dt, aes(x=factor(Depression), y=Age, fill=Depression)) + 
  geom_boxplot(alpha=0.7) + 
  labs(
    title="Correlation between age and depression",
    x="Depression",
    y="Age"
  )

ggplotly(p)
```

## Is academic pressure associated with depression?

There is a very strong correlation between academic pressure and depression (0.47), indicating that students experiencing academic pressure are more likely to experience depression.

```{r}
cor(dt$Academic.Pressure, dt$Depression)
```

```{r}
p <- ggplot(dt, aes(x=factor(Depression), y=Academic.Pressure, fill=Depression)) + 
  geom_boxplot(alpha=0.7) + 
  labs(
    title="Correlation between academic pressure and depression",
    x="Depression",
    y="Academic Pressure"
  )

ggplotly(p)
```

## Are students with higher work/study hours more likely to experience depression?

There is a correlation between work/study hours and depression (0.21), indicating that students working and studying more are more likely to experience depression.

```{r}
cor(dt$Work.Study.Hours, dt$Depression)
```

```{r}
p <- ggplot(dt, aes(x=factor(Depression), y=Work.Study.Hours, fill=Depression)) + 
  geom_boxplot(alpha=0.7) + 
  labs(
    title="Correlation between work/study hours and depression",
    x="Depression",
    y="Work/Study hours"
  )

ggplotly(p)
```

## Do dietary habits (e.g., healthy vs unhealthy) relate to depression?

The ANOVA results indicate a significant relationship between dietary habits and depression (p <2e-16). The results are illustrated in the bar plot below.

Perform ANOVA-test to examine the correlation
```{r}
anova_result <- aov(as.numeric(Depression) ~ as.factor(Dietary.Habits), data = dt)
```

```{r}
summary(anova_result)
```

```{r}
p <- ggplot(dt, aes(x=Dietary.Habits, fill=as.factor(Depression))) + 
  geom_bar(position = "stack") + 
  labs(
    title="Correlation between dietary habits and depression",
    x="Depression",
    y="Dietary habits",
    fill="Depression"
  )

ggplotly(p)
```


## Does financial stress correlate with depression?

There is a correlation between financial stress and depression (0.36), indicating that students experiencing financial stress are more likely to experience depression.

```{r}
cor(as.numeric(dt$Financial.Stress), as.numeric(dt$Depression))
```
remove NA values
```{r}
dt <- dt[!is.na(Financial.Stress)]
```

```{r}
cor(as.numeric(dt$Financial.Stress), as.numeric(dt$Depression))
```

```{r}
p <- ggplot(dt, aes(x=factor(Depression), y=Financial.Stress, fill=Depression)) + 
  geom_boxplot(alpha=0.7) + 
  labs(
    title="Correlation between financial stress and depression",
    x="Depression",
    y="Financial stress"
  )

ggplotly(p)
```

# Summary
The analysis showed that age, academic and financial pressure, work/study hours, and dietary habits are significantly associated with depression. 
No significant correlation was found between CGPA and depression.

